---
title: "producing input data for BACDAC"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is not an endorsement of the following tools, but a demonstration of how the input files may be produced.
Note, we have not tested these tools ourselves and cannot attest to the quality of the results.

# Basic Instructions

## ref alt counts using samtools

### snp positions 
You need a file with snp positions. `hetScore_dbSnp_20180418.tsv.gz` is our file with a list of 33629538 common snp positions. Other such files are available from a variety of sources. Our file is available to download at https://zenodo.org/records/18436153

The columns in `hetScore_dbSnp_20180418.tsv.gz` are: "chr", "pos", "id", "ref", "alt", "reffreq", "altfreq". 
To see the first 4 rows of data, skipping the comments and header line:

`zcat hetScore_dbSnp_20180418.tsv.gz | head | tail -n +7`

which should provide the following output:

```{Rmd}
 1  11012  rs544419019  C  G  0.9119  0.08806 
 1  11063  rs561109771  T  G  0.997   0.002995 
 1  13110  rs540538026  G  A  0.9732  0.02676 
 1  13116  rs62635286   T  G  0.903   0.09704 
```

### define some file names...
```{eval=FALSE}
# input the name of your files here.
bamFile=bamFile.bam
dbSnpPosFile=hetScore_dbSnp_20180418.tsv.gz
inputPosFile=inputPositionsList.tsv
pileupInit=pileup.tsv
pileupFinal=pileup_final.tsv
pileupTempFile=pileup_temp.tsv
inputHetFile=inputHetFile.tsv
```

### Make `$inputPosFile` 

Assuming your bam file is aligned to chromosomes labelled chr1,chr2,...,chrX,chrY.  
Extract the chr and position from your snp file.   
```{eval=FALSE}
zcat $dbSnpPosFile | tail -n +7 | awk '{if ($1==23) {test="X"} else if ($1==24) {test="Y"} else {test=$1} print "chr"test"\t"$2}' > $inputPosFile
```



### Samtools
Install samtools, some sudo commands might be needed

```{eval=FALSE}
wget https://github.com/samtools/samtools/releases/download/1.16.1/samtools-1.16.1.tar.bz2
tar -xvjf samtools-1.16.1.tar.bz2
cd samtools-1.16.1
./configure --prefix=/usr/local
make  
make install
```


### Running samtools
You can then use this as input to samtools mpileup to create a pileup file. The early commands are telling the program what to output, base quality, alignment quality.\  
Can limit depth (-d) in this case as we are getting REF/ALT balance and not depth itself.  
```{eval=FALSE}
samtools mpileup -a -A -q 10 -Q 15 -d 1000 --no-output-ins --no-output-ins --no-output-del --no-output-del --no-output-ends -l $inputPosFile -o $pileupInit $bamFile
# Can also do by chromosome using (-r) where [chromNum] is, for example `chr1`  
samtools mpileup -a -A -q 10 -Q 15 -d 1000 --no-output-ins --no-output-ins --no-output-del --no-output-del --no-output-ends -r [chromNum] -l $inputPosFile -o [pileupFileChr] $bamFile
```

### Conversion of samtools output
Convert the output into a more easily digestible form:
```{eval=FALSE}
samtools mpileup -a -A -q 10 -Q 15 -d 1000 --no-output-ins --no-output-ins --no-output-del --no-output-del --no-output-ends -l $inputPosFile -o $pileupInit $bamFile
# Can also do  by chromosome using (-r) where [chromNum] is, for example 'chr1'
samtools mpileup -a -A -q 10 -Q 15 -d 1000 --no-output-ins --no-output-ins --no-output-del --no-output-del --no-output-ends -r [chromNum] -l $inputPosFile -o [pileupFileChr] $bamFile
```

`$pileupFinal` can then be converted to the format needed

Here we process a pileup file to count base occurrences, format the output, and merge it with a database SNP file. It uses nested awk commands to count A, C, G, T, deletions, and refskip characters, then creates a final CSV or TSV? file with header "CHROM,POS,REF,ALT,COV,A,C,G,T,DEL,REFSKIP". 

```{eval=FALSE}
# 1. Process pileup file to count bases and deletions
cat $pileupInit | \
awk '{ p=$5; nA=gsub("[Aa]","",p); print $1,$2,$4,$5,nA; }' | \   # return how many A's are in the string
awk '{ p=$4; nC=gsub("[Cc]","",p); print $1,$2,$3,$4,$5,nC; }' | \
awk '{ p=$4; nG=gsub("[Gg]","",p); print $1,$2,$3,$4,$5,$6,nG; }' | \
awk '{ p=$4; nT=gsub("[Tt]","",p); print $1,$2,$3,$4,$5,$6,$7,nT; }' | \
awk '{ p=$4; nD=gsub("[*]","",p); print $1,$2,$3,$4,$5,$6,$7,$8,nD; }' | \
awk 'BEGIN {OFS=",";}{ p=$4; nR=gsub("[<>]","",p); print $3,$5,$6,$7,$8,$9,nR; }' \
> $pileupTempFile

# 2. Add Header to final output
echo "CHROM,POS,REF,ALT,COV,A,C,G,T,DEL,REFSKIP" > $pileupFinal

# 3. Merge processed pileup with database SNP file
paste -d',' \
  <(zcat $dbSnpPosFile | tail -n +7 | awk 'BEGIN {OFS=","};{print $1,$2,$4,$5 }') \
  <(cat $pileupTempFile) \
  >> $pileupFinal
```

Then extract specific allele counts (A, C, G, or T) from a genomic pileup file and reformat them into a 4-column summary file. 
It specifically identifies which count belongs to the "Reference" allele and which belongs to the "Alternate" allele for every genomic position.

```{eval=FALSE}
sw=( A:C:G:T )                           # Define a variable containing the four DNA bases separated by colons.
echo "chr,pos,ref,alt" > $inputHetFile  # Initialize the output file [inputHetFile] with a header row.
tail -n +2 $pileupFinal | awk -v sw="${sw[*]}" 'BEGIN {FS=","};BEGIN {OFS="\t"};{ 
  n=split(sw, AR, ":");   # Splits the string A:C:G:T into an array AR where AR[1]=A, AR[2]=C, etc.
  for(i=1;i<=4;i++) {     # Iterates through the four bases
    # Ref Allele Check: If the base in column 3 ($3) matches one of the bases in the array, it grabs the corresponding count from columns 6 through 9 (e.g., if $3 is 'A', it takes the value from $6).
    # Alt Allele Check: If the base in column 4 ($4) matches, it likewise grabs the count for that base.
    if($3==AR[i]) {
      rOut=$( 5 + i ); 
    }
    if($4==AR[i]) {
      aOut=$( 5 + i )
    }
  }
  print $1,$2,rOut,aOut    # Outputs the chromosome, position, and the extracted Ref/Alt counts
}' >> $inputHetFile
```


The resulting `$inputHetFile` will contain:  

1. Chromosome  
2. Position  
3. Count for the Reference Allele  
4. Count for the Alternate Allele  



## binned read count data and segmentation using HHMcopy

### HMMcopy_utils
The necessary input files for HMMcopy come from HMMcopy_utils.   

```{eval=FALSE}
git clone https://github.com/shahcompbio/hmmcopy_utils.git
cd hmmcopy_util  
cmake .  
make  
```

There are three scripts in hmmcopy_utils/bin: readCounter, gcCounter, mapCounter

* `readCounter` produces the coverage wig from bam
* `gcCounter` produces the gc wig from reference genome fasta
* `mapCounter` produces the mappability wig from a mappability reference file

### readCounter
set window (-w) to 30000 and limit to primary chromosomes

```{eval=FALSE}
# need to specify [filename]
hmmcopy_utils/bin/readCounter -w 30000 -c chr1,chr2,chr3,chr4,chr5,chr6,chr7,chr8,chr9,chr10,chr11,chr12,chr13,chr14,chr15,chr16,chr17,chr18,chr19,chr20,chr21,chr22,chrX,chrY` `[fileName].bam > [fileName].30kb.wig
```

### gcCounter
set window (-w) to 30000 and limit to primary chromosomes  
Use the hg38 genome.

```{eval=FALSE}
hmmcopy_utils/bin/gcCounter -w 30000 -c chr1,chr2,chr3,chr4,chr5,chr6,chr7,chr8,chr9,chr10,chr11,chr12,chr13,chr14,chr15,chr16,chr17,chr18,chr19,chr20,chr21,chr22,chrX,chrY` `[referenceGenomeFasta] > gc.hg38.30kb.wig
```


### mapCounter
First download a reference file, the 36bp Umap multiple mapping from UCSC

```{eval=FALSE}
wget http://hgdownload.soe.ucsc.edu/gbdb/hg38/hoffmanMappability/k36.Umap.MultiTrackMappability.bw
hmmcopy_utils/bin/mapCounter -w 30000 -c chr1,chr2,chr3,chr4,chr5,chr6,chr7,chr8,chr9,chr10,chr11,chr12,chr13,chr14,chr15,chr16,chr17,chr18,chr19,chr20,chr21,chr22,chrX,chrY` `k36.Umap.MultiTrackMappability.bw > map.hg38.k36.wig
```

These commands can also be run for 100Kb 
The user creates the gc/map files once. Then creates the read depth files for each bam.
 
Now use those files to run HMMCopy in R to produce the necessary files for BACDAC

### HMMCopy
```{r eval=FALSE}
library(HMMcopy)
rfile <- "[fileName].30kb.wig"
gfile <-"gc.hg38.wig"
mfile <- "map.hg38.k36.wig"
covIn <- wigsToRangedData(rfile, gfile, mfile)
covNorm <- correctReadcount(covIn)
# resort to order
covNorm <- covNorm[order(as.integer(factor(covNorm$chr,levels=paste0("chr",c(1:22,"X","Y"))))),]
# refactor the chromosome column
covNorm$chr <- factor(covNorm$chr,levels=paste0("chr",c(1:22,"X","Y")))
```

The columns of covNorm are: chr, start, end, reads, gc, map, valid, ideal, cor.gc cor.map, copy

The "valid" column is the mask.

HMMcopy has a built in program for segmentation also, though it tends to overcall. But primarily you want the segments. Could be better for using on the larger of the two windows, 100Kb.

For segmentation data

```{r eval=FALSE}
default_param <- HMMsegment(covNorm, getparam = TRUE)
longseg_param <- default_param
longseg_param$e <- 0.999999999999999
longseg_param$strength <- 1e30
longseg_segments <- HMMsegment(covNorm, longseg_param, verbose = FALSE)
```

The output here has `longseg_segments$segs` which gives how it has divided up the genome





